---
title: "Tumour microenvironment cell estimation (ConsensusTME)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  workflowr::wflow_html:
    toc_depth: 4
editor_options:
  chunk_output_type: console
---


```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


## ConsensusTME

```{r import}
library(RSQLite)
library(ConsensusTME)

# ConsensusTME::cancerAll # BRCA
```

The ConsensusTME package provides a integrative tool for R that uses a consensus approach to generating cancer specific signatures for multiple cell types found within the tumour microenvironment.

These consensus gene sets are then used within a ssGSEA framework to provide normalised enrichment scores for each of the cell types representing the relative abundance of cell types across multiple samples.

It is important to note that results generated in this way will provide quantification of cell types that are relative across samples rather than across cell types.

 returns estimation of cell type abundance for each sample in the bulk tumour gene expression matrix.
 Immune Score represents the overall level of immune infiltration for each sample.

```{r}
# sample_info table
sample_info = read.table("/hpcnfs/scratch/DIMA/piva/casper_ss_ma/data/samples_metadata.csv", header=T, sep = ',', row.names = 1)
rownames(sample_info) <- sample_info$samplename
sample_info$condition <- as.factor(sample_info$condition)
sample_info$batch <- as.factor(sample_info$batches)


# annotation
load('/hpcnfs/scratch/DIMA/piva/casper_ss_ma/data/data.Rdata')
#head(data) # TPM counts

# convert ENS gene id to Gene symbol
# gene symbol x patients # Matrix is a numerical matrix with gene symbol as rownames
matrix=data
genes=sub("\\..*", "", rownames(matrix))
length(genes)
length(unique(genes))

gene_map <- setNames(annotation$GeneSymbol, annotation$Gene)
symbols <- gene_map[genes]
sum(is.na(symbols))
sum(duplicated(symbols[!is.na(symbols)]))

summary_df <- data.frame(
  Total = length(symbols),
  Unique = length(unique(symbols)),
  NAs = sum(is.na(symbols)),
  Duplicated = sum(duplicated(symbols[!is.na(symbols)])),
  Empty = sum(symbols == "" | symbols == " ", na.rm = TRUE)
)
print(summary_df)

symbols <- make.unique(as.character(symbols))
matrix=apply(matrix, 2, as.numeric)
rownames(matrix)=symbols

matrix=matrix[!is.na(rownames(matrix)), ]
matrix=matrix[rownames(matrix)!=" ", ]
matrix=matrix[rownames(matrix)!="", ]

matrix[1:10,1:10]

```


```{r}
consensus=ConsensusTME::consensusTMEAnalysis(matrix, cancerType = 'BRCA', )

```

```{r}

consensus=t(consensus)
consensus=merge(consensus, sample_info, by = "row.names", all = TRUE )

```


# Correlation between TME enrichment and aneuploidy score (from CaSPeR)
```{r, fig.width=20, fig.height=25, out.width="100%"}
library(ggplot2)
library(ggpubr)    # For stat_cor()
library(patchwork) # For layout

target_var <- "aneuploidy_score"
feature_vars = colnames(consensus)[2:20]
df=consensus

# Create scatterplots with correlation
plots <- lapply(feature_vars, function(var) {
  ggplot(df, aes_string(x = var, y = target_var, color = "condition")) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top", size = 3.5) +
    ggtitle(var) +
    theme_minimal(base_size = 10)
})

# Combine into a grid
wrap_plots(plots, ncol = 4)


```

# Correlation between TME enrichment and CD82 gene expression

```{r, fig.width=20, fig.height=25, out.width="100%"}


target_var <- "ENSG00000085117"
feature_vars = colnames(consensus)[2:20]
df=consensus

# Create scatterplots with correlation
plots <- lapply(feature_vars, function(var) {
  ggplot(df, aes_string(x = var, y = target_var, color = "condition")) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top", size = 3.5) +
    ggtitle(var) +
    theme_minimal(base_size = 10)
})

# Combine into a grid
wrap_plots(plots, ncol = 4)

```


# Boxplot of TME enrichment across conditions

```{r, fig.width=20, fig.height=25, out.width="100%"}
df=consensus
df=df[!is.na(df$ENSG00000085117), ]
df$CD82_levels=ifelse(df$ENSG00000085117>median(df$ENSG00000085117), 'high-CD82', 'low-CD82')
df$aneuploidy_score_levels=ifelse(df$aneuploidy_score>median(df$aneuploidy_score), 'high-aneuploidy_score', 'low-aneuploidy_score')


condition_col <- "condition"
boxplot_vars = colnames(consensus)[2:20]

plots <- lapply(boxplot_vars, function(var) {
  ggplot(df, aes_string(x = condition_col, y = var, fill = condition_col)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.4, size = 0.8) +
    ggtitle(var) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
})
wrap_plots(plots, ncol = 4)


```

# Boxplot of TME enrichment across CD82 gene levels (stratified by median)

```{r, fig.width=20, fig.height=25, out.width="100%"}
condition_col <- "CD82_levels"
boxplot_vars = colnames(consensus)[2:20]

plots <- lapply(boxplot_vars, function(var) {
  ggplot(df, aes_string(x = condition_col, y = var, fill = condition_col)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.4, size = 0.8) +
    ggtitle(var) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
})
wrap_plots(plots, ncol = 4)


```

# Boxplot of TME enrichment across aneuploidy score levels (stratified by median)

```{r, fig.width=20, fig.height=25, out.width="100%"}
condition_col <- "aneuploidy_score_levels"
boxplot_vars = colnames(consensus)[2:20]

plots <- lapply(boxplot_vars, function(var) {
  ggplot(df, aes_string(x = condition_col, y = var, fill = condition_col)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.4, size = 0.8) +
    ggtitle(var) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
})
wrap_plots(plots, ncol = 4)

```
# Boxplot of TME enrichment across CD82 expression categories (by IHC) in TNBC

```{r, fig.width=20, fig.height=25, out.width="100%"}
condition_col <-"CATEGORY.1." # "CATEGORY.1." #"CD82_levels"
boxplot_vars = colnames(consensus)[2:20]

plots <- lapply(boxplot_vars, function(var) {
  ggplot(df[df$condition=="TNBC",], aes_string(x = condition_col, y = var, fill = condition_col)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.4, size = 0.8) +
    ggtitle(var) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
})
wrap_plots(plots, ncol = 4)


```

# Boxplot of TME enrichment across aneuploidy score levels in TNBC

```{r, fig.width=20, fig.height=25, out.width="100%"}
condition_col <- "aneuploidy_score_levels"
boxplot_vars = colnames(consensus)[2:20]

plots <- lapply(boxplot_vars, function(var) {
  ggplot(df[df$condition=="TNBC",], aes_string(x = condition_col, y = var, fill = condition_col)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.4, size = 0.8) +
    ggtitle(var) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
})
wrap_plots(plots, ncol = 4)

```


```{r, fig.width=20, fig.height=25, out.width="100%"}
condition_col <- "aneuploidy_score_levels"
boxplot_vars = colnames(consensus)[2:20]

plots <- lapply(boxplot_vars, function(var) {
  ggplot(df, aes_string(x = condition_col, y = var, fill = 'condition')) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.4, size = 0.8) +
    ggtitle(var) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
})
wrap_plots(plots, ncol = 4)

```

```{r}
#write.csv(consensus, '/hpcnfs/scratch/DIMA/piva/casper_ss_ma/data/consensusTME.csv')
```


